White Dwarf Mergers on Adaptive Meshes\\ II. Nuclear Reactions

### 2017-09-09 papers/wdmerger_II/paper.tex
Remove discussion of the dual-energy parameter
This has no serious effect on the outcome, so it ends up being a distraction.

Next is \aproxthirteen\ \citep{timmes:1999,timmes:2000}. This includes
all of the isotopes of \isoseven, and all of the $\alpha$-chain particles between
silicon and nickel (${}^{32}$S, ${}^{36}$Ar, ${}^{40}$Ca, ${}^{44}$Ti, ${}^{48}$Cr,
and ${}^{52}$Fe). This network was used by \citet{hawley:2012} and \citet{raskin:2010}.
\citet{loren-aguilar:2010} and \citet{garcia-senz:2013} used a very similar network
that included additionally ${}^{60}$Zn. The \aproxnineteen\ network \citep{timmes:1999}
builds on \aproxthirteen\ by including isotopes for hydrogen burning and explicit
tracking of photodisintegration into ${}^{54}$Fe. This network was used by
\citet{kushnir:2013}, \citet{kushnir:2014}, \citet{rosswog:2009} for their
calculations with FLASH, and \citet{papish:2015}. Finally we will also
consider \aproxtwentyone, which includes all of the above plus ${}^{56}$Cr
and ${}^{56}$Fe and related reaction links. The primary virtue of using
the latter two networks is that they allow us to track changes away from
$Y_e = 0.5$.

@@

One other point worth noting for the coupling of the reactions
and the hydrodynamics relates to the dual-energy formalism used
by \castro\ and many other hydrodynamics codes \cite{bryan:1995}.
\castro\ evolves separately equations for the total energy $(\rho E)$ and
for the internal energy $(\rho e)$. The former is conservative while the latter
is not, so for accurate hydrodynamics we prefer to use the total
energy when possible and to use the evolved internal energy variable
only in situations where the kinetic energy dominates the contribution
to the total energy. Indeed, for the cosmological purpose for which
this was originally developed, \cite{ENZO} choose a value $\eta_2 = 0.1$
so that $e$ is only updated to be equal to $E - \mathbf{v}^2/2$ if
$e > \eta_2\, E$. In other cases the evolved value of the internal
energy variable is preserved. However, this choice is somewhat unsafe
for our application, because the ultimate cause of the nuclear
burning in a white dwarf collision is the rapid conversion of
kinetic energy from the white dwarf bulk motion to thermal energy
as the white dwarfs slam into each other. Keeping $\eta_2$ large
prevents this from happening, and consequently the temperature
will never reach values high enough to generate significant
amounts of nickel. For this paper we choose $\eta_2 = 10^{-4}$,
which is low enough for the correct conversion of kinetic energy
but not so low that we need to be concerned about roundoff issues
caused by subtracting kinetic from total energy. This is also
the value that was used by \cite{hawley:2012}.

@@

\castro\ uses the dual energy formalism of \cite{bryan:1995,ENZO}, where we evolve equations
for both the total energy $E$ and the internal energy $e$, using one or the other to
determine the pressure in the hydrodynamics step (if we are using the total energy, we
compute the corresponding internal energy by subtracting the kinetic energy $K$). The switching
parameter is $\eta_1 = 10^{-3}$, and if $(E - K) / E < \eta_1$, we use the internal energy variable $e$
in the equation of state call that obtains the pressure. Otherwise we use $(E - K)$. The
idea behind this is that in high Mach number flows where the kinetic energy dominates the
total energy, we do not want to risk generating a negative estimate of the internal energy
due to roundoff error. Similarly, at regular intervals when we enforce consistency of the
internal energy variable (which occurs a few times per timestep), we sync the internal energy
variable so that it is equal to $E - K$, skipping this step only if $(E - K) / E < \eta_2$.
We adopt by default $\eta_2 = 10^{-4}$. This choice can matter much more for reacting
hydrodynamics than non-reacting hydrodynamics. For example, a collision can only lead to a
detonation if there is sufficient conversion of kinetic to thermal energy. Obviously
this happens physically, but if the parameter $\eta_2$ is set high enough then it could
in principle interfere by not allowing the simulation to efficiently convert kinetic energy
to thermal energy. (This does not end up mattering much for our simulations because the
kinetic energy does not strongly dominate the internal energy for colliding WDs.) And on
a similar note, when using this formalism it is ambiguous which variable to use ($(E - K)$
or $e$) when calculating the initial temperature that goes into the reaction step. We
have introduced a new parameter $\eta_3$ that uses the same logic as the first two: if
$(E - K) / E < \eta_3$, we use $(E - K)$ in the equation of state call that obtains the
temperature; otherwise, we use $e$. However, unlike the other two, the \castro\ default
is $\eta_3 = 1.0$, meaning that we normally use the internal energy variable.

@@

We tested a few other parameters that ended up having no serious effect on the outcome.
The dual energy parameter $\eta_2$ described in \autoref{sec:problemsetup} can have a
minor effect on the nickel production, yielding variations on the order of a few percent,
but there is no clear pattern. The dual energy parameter $\eta_3$ has no effect on the
outcome: the answer is the same for $\eta_3 = 0$ and $\eta_3 = 1$, which makes sense
because all of the reacting regions have a low enough kinetic energy relative to their
total energy such that the internal energy variable is always kept in sync with $(E - K)$.
(We did not look at $\eta_1$ because that has a direct effect only on the hydrodynamics,
not the reactions.) We also tested whether our choice for the temperature floor mattered
by lowering it to $10^6\ \text{K}$ and $10^5\ \text{K}$; this had a minor effect on nickel
production, again on the order of a few percent. It does allow parts of the WDs to
become colder as they move, due to numerical errors in the advecting flow (the Helmholtz
equation of state is very insensitive to temperature, so even small changes in the internal
energy can create very large changes in the implied temperature), but the amount of energy
injected by the collision is so large that the final result is largely insensitive to whatever
the initial temperature of the
WDs were. We also checked whether there was any consequence of our choice to disable the
burning for small enough $\rho$ and $T$. Lowering the minimum $\rho$ for reactions to
$10^0\ \text{g cm}^{-3}$ from $10^6\ \text{g cm}^{-3}$ had a negligible effect,
and varying the minimum $T$ for reactions from $2 \times 10^{7}\ \text{K}$ to
$4 \times 10^{8}\ \text{K}$ also had no meaningful effect.

>>>>

Next is \aproxthirteen\ \citep{timmes:1999,timmes:2000}. This includes
all of the isotopes of \isoseven, and all of the $\alpha$-chain particles between
silicon and nickel (${}^{32}$S, ${}^{36}$Ar, ${}^{40}$Ca, ${}^{44}$Ti, ${}^{48}$Cr,
and ${}^{52}$Fe). This network was used by \citet{hawley:2012} and \citet{raskin:2010}.
\citet{loren-aguilar:2010} and \citet{garcia-senz:2013} used a very similar network
that included additionally ${}^{60}$Zn. The \aproxnineteen\ network \citep{timmes:1999}
builds on \aproxthirteen\ by including isotopes for hydrogen burning and explicit
tracking of photodisintegration into ${}^{54}$Fe. This network was used by
\citet{kushnir:2013}, \citet{kushnir:2014}, and \citet{rosswog:2009} for their
calculations with FLASH, and \citet{papish:2015}. Finally we will also
consider \aproxtwentyone, which includes all of the above plus ${}^{56}$Cr
and ${}^{56}$Fe and related reaction links. The primary virtue of using
the latter two networks is that they allow us to track changes away from
$Y_e = 0.5$.

@@

@@

@@

We tested a few other parameters that ended up having no serious effect on the outcome.
The dual energy hydrodynamic formalism (discussed in Paper I) has parameters that determine
whether we obtain the internal energy from the total energy or from a separately advected
internal energy variable. This is usually employed when calculating the pressure during
a hydrodynamics update, and we have also added to \castro\ a similar switch when calculating
the initial temperature for a burn. We find that the value of the dual energy parameter has
only a minor effect on the nickel production, yielding variations on the order of a few percent
with no systematic trend. We also tested whether our choice for the temperature floor mattered
by lowering it to $10^6\ \text{K}$ and $10^5\ \text{K}$; this had a minor effect on nickel
production, again on the order of a few percent. It does allow parts of the WDs to
become colder as they move, due to numerical errors in the advecting flow (the Helmholtz
equation of state is very insensitive to temperature, so even small changes in the internal
energy can create very large changes in the implied temperature), but the amount of energy
injected by the collision is so large that the final result is largely insensitive to whatever
the initial temperature of the
WDs were. We also checked whether there was any consequence of our choice to disable the
burning for small enough $\rho$ and $T$. Lowering the minimum $\rho$ for reactions to
$10^0\ \text{g cm}^{-3}$ from $10^6\ \text{g cm}^{-3}$ had a negligible effect,
and varying the minimum $T$ for reactions from $2 \times 10^{7}\ \text{K}$ to
$4 \times 10^{8}\ \text{K}$ also had no meaningful effect.
